Bayesian model in JAGS

### 2023-01-31 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
update tutorials

>>>>

Please open Jags_model.R for the whole R script.
\
\
$k=1,2,3$ represent each socioeconomic groups, $j=1,...,9$ represent each age groups, and $i=1,...,108$ represent each month.
\
\
Hierarchical Bayesian model are specified in jags as follows:
```{r, eval=FALSE}
model {
for(k in 1:n.group){  
  for (j in 1:n.age){
      for (i in 1:n.date) {

 y[i,j,k] ~ dnegbin(prob[i,j,k],r) 
 # negative binomial
  prob[i,j,k]<- r/(r+lambda[i,j,k])  ## likelihood 
```

## R code (JAGS code): expected respiratory hospitalizations

```{r, eval=FALSE}
# expected respiratory hospitalizations lambda
    lambda[i,j,k] <- rd0[j,k]+ # baseline hosp
exp(epi[epi.year[i]]) + # year to year variation
# epi.year[i] is an input indicator for the calendar year
# epi.year[i] = 1,2,...,9
exp(delta[month[i]]) + # monthly variations
# month[i] is an input indicator for the epi month
# month[i] = 1,2,...,12 (1 for July and 12 for June)
rsv[i,j,k]*rd2[j,k] + # RSV attributable hosp
flu[i,j,k]*rd1[epi.year[i],j,k] # flu attributable hosp
      }

# baseline hospitalizations 
#(must be greater than or equal to 0)
  rd0[j,k] <- exp(beta0[j,k])
  beta0[j,k]~ dnorm(mu0,tau0) 
# hierarchical structure: shared mean and variance
```

## The coefficient of flu in each age and SES group

```{r, eval=FALSE}
 for (p in 1:n.year) { 
# epi-year effects for antigen shift
# coefficient of influenza-associated respiratory 
# hospitalization varies annually 
    rd1[p,j,k] <- exp(beta1[p,j,k]) 
    # ensure positive coefs
    beta1[p,j,k] ~ dnorm(beta1_mean[p,j,k],tau.flu)
    # shared variance
    # the mean of the coefficient is 
# a multiplicative combination of 
    beta1_mean[p,j,k] <- gamma_flu[k]+ omega_flu[j]+ xi_flu[p]
    # SES effects gamma_flu[k]
    # age effects omega_flu[j]
    # epi-year (antigen) xi_flu[p]
    }
```

## The coefficient of RSV in each age and SES group

```{r, eval=FALSE}
 # this coefficient depends on SES and age 
    rd2[j,k] <- exp(beta2[j,k]) # ensure positive
    beta2[j,k] ~ dnorm(beta2_mean[j,k],tau.rsv)
    beta2_mean[j,k]<- gamma[k]+omega[j]
   # SES effects gamma[k]
  # age effects omega[j]
```

## hyperparameters

```{r, eval=FALSE}
# n.year = 1,2,...,9
for (p in 1:n.year) { 
epi[p] ~ dnorm(0, tau.epi) 
# prior for yearly variation of baseline hosp
xi_flu[p] ~ dnorm(0,tau7)}
# prior for flu annual variation (antigen shift)

  for (m in 1:12){
    delta[m] ~ dnorm(0,disp.m)
  }
# prior for monthly variation of baseline hosp
```

## hyperparameters (continued)

```{r, eval=FALSE}
for(k in 1:n.group){
gamma_flu[k] ~ dnorm(0,tau3)
gamma[k] ~ dnorm(0, tau4)
}
# prior for SES effects of flu and RSV

for(j in 1:n.age){
omega[j] ~ dnorm(0, tau5)
omega_flu[j] ~ dnorm(0,tau6)
}
# prior for age effects of flu and RSV
```

## Priors
```{r, eval=FALSE}
 r ~ dunif(0,250) 
#r > 0 serves as an overdispersion parameter
  mu0 ~ dnorm(0,0.0001) # 0.0001 = 1/variance
# In jags, dnorm specify precision

# the conjugate priors for variance of normal distribution
# is an inverse Gamma distribution
# In jags, we put priors on precision
  tau0 ~ dgamma(0.01, 0.01)
  tau1 ~ dgamma(0.01, 0.01)
  tau2 ~ dgamma(0.01, 0.01)
  tau3 ~ dgamma(0.01, 0.01)
  tau4 ~ dgamma(0.01, 0.01)
  tau5 ~ dgamma(0.01, 0.01)
  tau6 ~ dgamma(0.01, 0.01)
  tau7 ~ dgamma(0.01, 0.01)
  tau8 ~ dgamma(0.01, 0.01)
  tau.epi ~ dgamma(0.01, 0.01)
  tau.flu ~ dgamma(0.01, 0.01)
  tau.rsv ~ dgamma(0.01, 0.01)
  disp.m ~ dgamma(0.01, 0.01)
```

## Clarification for input data

**Note: we coded as rsv[i,j,k] and flu[i,j,k] for the purpose of a clear dimension for multiplication.**  This is a 1 to 1 multiplication, not a matrix multiplication. rsv[i,j,k] are the same for j=1,...,9, representing ICD-9 recorded RSV hospitalizations in children under 2 in each SES group in each month. flu[i,j,k] are the same for j=1,...,9, representing the total ICD-9 recorded flu hospitalizations in the entire age spectrum in each SES group in each month.

\
\

In the equation, we simplify them to $RSV_{ik}$ and $Flu_{ik}$.

## Specify input data

```{r, eval=FALSE}
epi.year <- as.factor(rep(1:9, each=12))
# indicator for calendar year
month <- rep(1:12,9)
# indicator for epi month (1 for July and 12 for June)

dataset <- list('y' = y_income_whole, 
# respiratory hospitalizations in each age and  
# SES group in each month
"rsv"=rsv_income_whole,
# ICD-9 recorded RSV hospitalizations   
# in children under 2 in each SES group in each month
"flu"=flu_income_whole, 
# the total ICD-9 recorded flu hospitalizations in 
# the entire age spectrum in each SES group in each month 
'epi.year'=epi.year,"month"=month,
n.age=9,n.date=108,n.year=9,"n.group"=3) 
# number of age/SES groups; number of months and years
```

## Posterior sampling

```{r, eval=FALSE}
jags_post <- jags.model(
  textConnection(model_string), 
  data = dataset,n.chains = 3)
# specify the number of chains

update(jags_post, n.iter=82500) # burn-in period

rsv_resp <- coda.samples(jags_post,
                         variable.names=
  c("rd2","lambda","rd1", "rd0","epi","delta"),
                      thin = 10,n.iter = 12500)
# posterior samples
```

## Estimate the RSV-attributable hospitalizations in each age and SES group

```{r, eval=FALSE}
## Remember to check convergence first
## use: gelman.diag() see 3_JAGS_Bayesian_model
## Calculate RSV attributable respiratory 
## hospitalization incidence and percent
  post1 <- as.data.frame(as.matrix(rsv_resp[[1]]))
  post2 <- as.data.frame(as.matrix(rsv_resp[[2]]))
  post3 <- as.data.frame(as.matrix(rsv_resp[[3]]))
  post <- bind_rows(post1,post2,post3)
## A total of 3750 posterior samples
  lambda <- post[, grep("lambda", colnames(post), fixed=T)]
  rd2.resp <- post[, grep("rd2[", colnames(post), fixed=T)]

rsv_count <- array(data = NA,dim = c(3750,108,9,3))
for (i in 1:108) {
  for (j in 1:9) {
    for (k in 1:3) {
      rsv_count[,i,j,k] <- rsv_income[i,j,k]*rd2.resp[,j+9*(k-1)]}}}
## posterior samples of RSV-attributable hospitalizations
## in each SES and age group in each month
```
